---
title: "Final Project-Presentation"
author: "Thuy Minh Nhan Do"
date: "2025-11-12"
output: slidy_presentation
---
```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel) # For labels in bubble chart
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
majors <- read_csv("all-ages.csv")
```

## Introduction 
- College majors greatly shape graduates’ income, job stability, and career direction.  
- While many assume STEM majors always earn the most, the data reveal a more complex story.  
- This project explores how salaries, unemployment, and major size interact to show real-world outcomes across fields.

## 1. Dataset

**Dataset:** FiveThirtyEight “College Majors”  
**Source:** [GitHub](https://github.com/fivethirtyeight/data/tree/master/college-majors)

Columns include:

  - `Major`  
  
  - `Major_category`  
  
  - `Total`, `Employed`, `Employed_full_time_year_round`  
  
  - `Unemployed`, `Unemployment_rate`  
  
  - `Median`

## Why It’s Interesting
- Compare **assumptions vs. reality**: Are STEM majors always highest-paid?  
- Examine **employment outcomes** across majors  
- Identify majors with **unexpectedly high/low salaries or unemployment rates**  
- Highlight **career insights** for future graduates  

## Project Focus
We focus on:  

1. **Salary trends:** Which majors pay well? Any exceptions? 

2. **Employment outcomes:** High-paying majors with high unemployment?  

3. **Major size effect:** Does the number of graduates impact outcomes?  

4. **Category comparison:** STEM vs. non-STEM fields  

## Research Questions
- Which majors are unexpectedly high-paying or low-paying compared to their category stereotype?
- Are there majors with high unemployment despite high median salary, or vice versa?
- Which majors have the top salaries overall?
- Do larger graduating classes in a major correlate with better employment outcomes?

## Salary by Major Category
```{r}
majors %>%
group_by(Major_category) %>%
summarise(avg_salary = mean(Median)) %>%
  
ggplot(aes(x = reorder(Major_category, avg_salary), y = avg_salary, fill = avg_salary)) +
geom_col() +
coord_flip() +
labs(title = "Average Median Salary by Major Category",
x = "Major Category",
y = "Average Median Salary (USD)") +
theme_minimal() +
theme(legend.position = "none")
```

- STEM majors generally highest-paid

- Business and Healthcare show strong salaries

- Humanities and Arts lower on average

## Median Salary vs. Unemployment Rate
```{r}
majors %>%
  ggplot(aes(x=Median, y=Unemployment_rate, size=Total, color=Unemployment_rate)) +
  geom_point(alpha=0.7) +
  scale_size(range=c(2, 12), guide=guide_legend(title="Total Graduates")) +
  scale_color_gradient(low="lightblue", high="darkblue", guide=guide_colorbar(title="Unemployment Rate")) +
  scale_x_continuous(labels=scales::dollar) +
  labs(
    x="Median Salary",
    y="Unemployment Rate",
    title="Median Salary vs. Unemployment Rate, Sized by Total Graduates"
  ) +
  theme_minimal()
```

## Top 10 highest-paying majors
```{r}
top10 <- majors %>%
arrange(desc(Median)) %>%
slice_head(n=10)

ggplot(top10, aes(x=reorder(Major, Median), y=Median)) +
geom_segment(aes(x=Major, xend=Major, y=0, yend=Median), color="gray") +
geom_point(color="darkred", size=4) +
coord_flip() +
labs(x="Major", y="Median Salary", title="Top 10 Highest-Paying Majors") +
theme_minimal()

```

- Engineering and Computer Science dominate

- Healthcare fields appear in top 10

- Significant pay gap between top majors

## Total graduates vs. unemployment rate
```{r}
majors %>%
ggplot(aes(x=Total, y=Unemployment_rate)) +
geom_point(alpha=0.6, color="purple") +
scale_x_log10() +
labs(x="Total Graduates", y="Unemployment Rate", title="Total Graduates vs. Unemployment Rate") +
theme_minimal()

```


