---
title: "Final Project-Presentation"
author: "Thuy Minh Nhan Do"
date: "2025-11-12"
output: slidy_presentation
---
```{r echo = F, message = F, warning= F}
library(tidyverse)
library(ggplot2)
library(scales)
library(ggrepel) # For labels in bubble chart
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
majors <- read_csv("all-ages.csv")
recent <- read_csv("recent-grads.csv")
women_stem <- read_csv("women-stem.csv")
```

## Introduction 
- College majors shape graduates’ income, job stability, and career trajectory.
- STEM majors are often assumed to earn the most, but the data reveal a more nuanced reality.
- This project explores salaries, unemployment, major size, gender, and income range across fields.

## 1. Dataset

**Dataset:** FiveThirtyEight “College Majors”  
**Source:** [GitHub](https://github.com/fivethirtyeight/data/tree/master/college-majors).

- **all-ages.csv**: Employment & salary for all workers by major.
- **recent-grads.csv**: Outcomes for recent graduates.
- **women-stem.csv**: Gender representation and salaries in STEM majors.

Columns include:

  - `Major`  
  
  - `Major_category`  
  
  - `Total`, `Employed`, `Employed_full_time_year_round`  
  
  - `Unemployed`, `Unemployment_rate`  
  
  - `Median`
  - `P25th`, `P75th`
  - `Men`, `Women`, `ShareWomen`

## Why It’s Interesting
- Compare **assumptions vs. reality**: Are STEM majors always highest-paid?
- Explore **gender disparities** in STEM majors.
- Examine **employment outcomes** across majors.  
- Identify majors with **unexpectedly high/low salaries or unemployment rates**.
- Investigate **salary ranges (P25–P75)** for income inequality insights.  

## Project Focus
We focus on:  

1. **Salary trends:** Which majors pay well? Any exceptions? 

2. **Employment outcomes:** High-paying majors with high unemployment?  

3. **Major size effect:** Does the number of graduates impact outcomes?  

4. **Category comparison:** STEM vs. non-STEM fields.

5. **Gender representation** in STEM.

6. **Income range** analysis.

7. **Unemployment comparison:** All workers vs. recent graduates.

## Research Questions
- Which major categories offer the highest median salaries?
- How does median salary relate to unemployment rate across majors?
- Which majors have the top median salaries overall?
- Does major size correlate with unemployment rate?
- How does salary range (P25–P75) differ across majors?
- How does the share of women in STEM relate to median salary?
- How does unemployment differ between all workers and recent graduates?

## Salary by Major Category
```{r}
majors %>%
group_by(Major_category) %>%
summarise(avg_salary = mean(Median)) %>%
  
ggplot(aes(x = reorder(Major_category, avg_salary), y = avg_salary, fill = avg_salary)) +
geom_col() +
coord_flip() +
labs(title = "Average Median Salary by Major Category",
x = "Major Category",
y = "Average Median Salary (USD)") +
theme_minimal() +
theme(legend.position = "none")
```

- STEM majors earn the highest overall salaries.
- Business and Healthcare also pay well.
- Humanities and Arts generally have lower earnings.

## Median Salary vs. Unemployment Rate
```{r}
majors %>%
  ggplot(aes(x=Median, y=Unemployment_rate, size=Total, color=Unemployment_rate)) +
  geom_point(alpha=0.7) +
  scale_size(range=c(2, 12), guide=guide_legend(title="Total Graduates"), labels = comma) +
  scale_color_gradient(low="lightblue", high="darkblue", guide=guide_colorbar(title="Unemployment Rate")) +
  scale_x_continuous(labels=scales::dollar) +
  labs(
    x="Median Salary",
    y="Unemployment Rate",
    title="Median Salary vs. Unemployment Rate, Sized by Total Graduates"
  ) +
  theme_minimal()
```

- Higher salaries usually mean lower unemployment, but some exceptions exist.
- Major size isn’t a guarantee of job security.

## Top 10 highest-paying majors
```{r}
top10 <- majors %>%
arrange(desc(Median)) %>%
slice_head(n = 10) %>%
mutate(Major = fct_reorder(Major, Median))

ggplot(top10, aes(x = Major, y = Median)) +
geom_segment(aes(x = Major, xend = Major, y = 0, yend = Median), color = "gray") +
geom_point(color = "darkred", size = 4) +
coord_flip() +
scale_y_continuous(labels = dollar) +
labs(title = "Top 10 Highest-Paying Majors",
x = "Major",
y = "Median Salary (USD)") +
theme_minimal()

```

- Engineering and Computer Science dominate the top-paying majors.
- Healthcare fields also appear in the top 10.
- Even within the top 10, Engineering-related majors stand out with noticeably higher median salaries.

## Total graduates vs. unemployment rate
```{r}
majors %>%
ggplot(aes(x=Total, y=Unemployment_rate)) +
geom_point(alpha=0.6, color="orange") +
scale_x_log10(labels = comma) +
labs(x="Total Graduates", y="Unemployment Rate", title="Total Graduates vs. Unemployment Rate") +
theme_minimal()

```

- Smaller majors show more variability in unemployment.
- Larger majors’unemployment rates are more consistent.
- Major size alone does not determine job security.

## Salary Range: P25–P75 by Major
```{r}
top15_median <- majors %>%
arrange(desc(Median)) %>%
slice_head(n = 15) %>%
mutate(Major = fct_reorder(Major, Median))

ggplot(top15_median, aes(x = Major, y = Median)) +
geom_point(size = 3, color = "darkred") +
geom_errorbar(aes(ymin = P25th, ymax = P75th), width = 0.2, color = "gray") +
coord_flip() +
scale_y_continuous(labels = dollar) +
labs(title = "Top 15 Majors: Median with P25–P75 Range",
x = "Major",
y = "Salary (USD)") +
theme_minimal()

```

- Shows median salary and P25–P75 range for top 15 majors.
- Technical/Engineering fields have the widest ranges.
- Highlights variability in earnings within each major.

##Median Salary vs Share of Women in STEM
```{r}
women_stem %>%
  ggplot(aes(x = ShareWomen, y = Median)) +
  geom_point(alpha = 0.7, color = "purple") +
  geom_smooth(
    method = "lm",
    se = TRUE,
    alpha = 0.15,       
    color = "black",
    fill = "purple"       
  ) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Share of Women vs Median Salary in STEM Majors",
    x = "Share of Women",
    y = "Median Salary (USD)"
  ) +
  theme_minimal()

```

- Plots share of women vs median salary in STEM majors.
- Shows a slight negative trend: higher female share often links to lower median salaries.
- Highlights potential gender pay disparities within STEM fields.

## Unemployment: All Workers vs Recent Graduates
```{r}
combined <- bind_rows(
majors %>% select(Unemployment_rate) %>% mutate(Group = "All Workers"),
recent %>% select(Unemployment_rate) %>% mutate(Group = "Recent Graduates")
)

ggplot(combined, aes(x = Unemployment_rate, color = Group, fill = Group)) +
geom_density(alpha = 0.3) +
labs(title = "Unemployment Rate Distribution: All Workers vs Recent Graduates",
x = "Unemployment Rate",
y = "Density") +
theme_minimal()


```

- Compares unemployment rate distributions for all workers vs recent graduates.
- Recent graduates tend to have slightly higher unemployment.
- Some fields remain consistently low or high regardless of experience.

## Summary & Key Insights

- Salary Trends: STEM, Business, and Healthcare pay the most; Arts & Humanities lower.

- Employment: Higher salaries often lower unemployment, size doesn't guarantee security.

- Top Majors: Engineering & CS lead; Healthcare also strong.

- Salary Range: Tech fields wide, Arts/Humanities narrow.

- Gender in STEM: Higher female share may correlate with lower salaries.

- Recent Grads: Slightly higher unemployment than all workers; some fields consistent.

